One-stage point transect distance sampling using iterated integrated nested Laplace approximations

### 2024-07-25 paper.tex
add inlabru reference + edit iinla section

\documentclass{stylefile16/statsoc}
\usepackage[a4paper]{geometry}   % otherwise statsoc class doesn't compile to A4 correctly
%\usepackage{setspace}
%\doublespacing

\usepackage[hidelinks]{hyperref} % auto detect ref type
\usepackage{natbib}
\setcitestyle{stylefile16/rss}

@@

% shortcuts
\newcommand{\bbeta}{\boldsymbol{\beta}}
\newcommand{\blambda}{\boldsymbol{\lambda}}
\newcommand{\T}{\intercal}
\newcommand{\bs}{\mathbf{s}}
\newcommand{\bS}{\mathbf{S}}
\newcommand{\bQ}{\mathbf{Q}}
\newcommand{\bSigma}{\boldsymbol{\Sigma}}
\newcommand{\bm}{\boldsymbol}  % bold maths symbols
\newcommand{\tl}{\tilde{\lambda}}   % thinned little lambda
\newcommand{\tL}{\tilde{\Lambda}}  % thinned big lambda

@@

\author[Andrew E. Seaton {\it et al.}]{Andrew E. Seaton}
\address{Centre for Research into Ecological \& Environmental Modelling and School of Mathematics \& Statistics, 
	University of St Andrews, 
	St Andrews,
	UK}
\email{aes22@st-andrews.ac.uk}

@@

The half-normal detection function depends on the strictly positive parameter $\sigma^2$. Since it is not possible to restrict the range of parameters to be strictly positive within \texttt{R-INLA} this model cannot be fitted using the standard INLA methodology.  Instead, we choose a log link and consider $g(\bs | \phi)$, where $\log\sigma^2 = \phi$. This is a non-linear function of $\phi$ which cannot be fitted using standard GLM methods.  To address this we use a novel approximate method based on a first-order Taylor approximation of the non-linear predictor.

@@

Let $\bar{p}_{\bm{v}}$ denote the INLA approximate posterior distribution of the model parameters given the linearised model configuration at linearisation point $\bm{v}$.  New linearisation points are found by considering the functional
\begin{equation}
	f(\bar{p}_{\bm{v}}) = \left\{ \argmax_{\bm{u}_i} \: \bar{p}_{\bm{v}}(u_i | \bm{Y}), \quad i = 1, \ldots, n_{\bm{u}} \right\}
\end{equation}
where $n_{\bm{u}}$ is the length of $\bm{u}$.  Each marginal posterior $\bar{p}_{\bm{v}} (u_i | \bm{Y})$ and their modes are estimated during the model fitting using the standard INLA method and so this functional can be evaluated at no extra cost for each fit of the linearised model.
We then seek a fixed point of this functional so that $\bm{u}_0 = f(\bar{p}_{\bm{u}_0})$.  This fixed point is identified via an iterative scheme. Given a current choice of linearisation point $\bm{v}$, we estimate $\bar{p}_{\bm{v}}$ and set $\bm{u}_* :=f(\bar{p}_v)$ to use as the new linearisation point.  We iterate this procedure until a fixed point is identified within a chosen tolerance.

In practice the point $\bm{u}_*$ may be problematic and result in a linearised model configuration that is challenging to fit.  Instead this point is considered only as an initial candidate point.  The next linearisation point is chosen by setting $\bm{u}_{\alpha} = (1 - \alpha)\bm{v} + \alpha \bm{u}_*$ and finding the value of $\alpha$ that minimises $ \lVert \eta(\bm{u}_{\alpha}) - \bar{\eta}_{\bm{v}} \rVert$.  This means that the new linearisation point remains close to a feasible region for fitting, assuming the current point $\bm{v}$ is reasonable.  In practice, we have found that this requires some care when choosing starting values for the initial model fit.  We estimate $\alpha$ using an approximate line search method that avoids many potentially expensive evaluations of the non-linear predictor.  This iterative method is implemented in the R package \texttt{inlabru} \citep{bachl_inlabru_2019} which is both a wrapper and an extension to \texttt{R-INLA}.

@@

\clearpage
\bibliographystyle{stylefile16/rss}
\bibliography{paper}

>>>>

\documentclass{statsoc}
\usepackage[a4paper]{geometry}   % otherwise statsoc class doesn't compile to A4 correctly
%\usepackage{setspace}
%\doublespacing

\usepackage[hidelinks]{hyperref} % auto detect ref type
\usepackage{natbib}
%\setcitestyle{rss}

@@

% new commands
\newcommand{\wt}[1]{\widetilde{#1}}
\newcommand{\ol}[1]{\overline{#1}}
\newcommand{\wh}[1]{\widehat{#1}}
\newcommand{\bbeta}{\boldsymbol{\beta}}
\newcommand{\blambda}{\boldsymbol{\lambda}}
\newcommand{\T}{\intercal}
\newcommand{\bs}{\mathbf{s}}
\newcommand{\bS}{\mathbf{S}}
\newcommand{\bQ}{\mathbf{Q}}
\newcommand{\bSigma}{\boldsymbol{\Sigma}}
\newcommand{\bm}{\boldsymbol}  % bold maths symbols
\newcommand{\tl}{\tilde{\lambda}}   % thinned little lambda
\newcommand{\tL}{\tilde{\Lambda}}  % thinned big lambda

@@

\author[Andrew E. Seaton {\it et al.}]{Andrew E. Seaton}
\address{School of Mathematics \& Statistics, 
	University of Glasgow, 
	Glasgow,
	UK}
\email{andrew.seaton.2@glasgow.ac.uk}

@@

The half-normal detection function depends on the strictly positive parameter $\sigma^2$. Since it is not possible to restrict the range of parameters to be strictly positive within \texttt{R-INLA} this model cannot be fitted using the standard INLA methodology.  Instead, we choose a log link and consider $g(\bs | \phi)$, where $\log\sigma^2 = \phi$. This is a non-linear function of $\phi$ which cannot be fitted using standard GLM methods.  To address this we use a novel approximate-method based on an iterative model fitting scheme. This method is implemented in the \texttt{R} package \texttt{inlabru} which is a wrapper for and an extension of the \texttt{R-INLA} package \citep{lindgren_inlabru_2024}.  Full details of the method can be found in the software paper, here we present a brief overview.

@@

Let $\bar{p}_{\bm{v}}$ denote the INLA approximate posterior distribution of the model parameters given the linearised model configuration at linearisation point $\bm{v}$.  New linearisation points are found by via the functional
\begin{equation}
	f(\ol{p}_{\bm{v}}) = (\wh{\bm{\theta}}_{\bm{v}},\wh{\bm{u}}_{\bm{v}})
\end{equation}
where
\begin{equation}
\wh{\bm{\theta}}_{\bm{v}} = \argmax_{\bm{\theta}} \ol{p}_{\bm{v}} ( \bm{\theta} | \bm{y}),
\end{equation}
the posterior mode for $\bm{\theta}$, and
\begin{equation}
\wh{\bm{u}}_{\bm{v}} = \argmax_{\bm{u}} \ol{p}_{\bm{v}} (\bm{u} | \bm{y}, \wh{\bm{\theta}}_{\bm{v}}),
\end{equation}
the joint conditional posterior mode for $\bm{u}$.  This functional defines a new linearisation point as the posterior mode for $\bm{\theta}$ under the linearised model configuration and the conditional mode for $\bm{u}$ given $\bm{\theta}$.  These posteriors are computed automatically during the model fitting using the standard INLA method and so this functional can be evaluated at no extra cost for each fit of the linearised model.

The `optimal' approximate model is defined as a fixed point of this functional.  Given a current choice of linearisation point $\bm{v}$, we estimate $\bar{p}_{\bm{v}}$ and set $\bm{u}_* :=f(\bar{p}_{\bm{v}})$ to use as the new linearisation point.  We iterate this procedure until a fixed point $\bm{v}^*$ is identified within a chosen tolerance such that $\bm{v}^* = f(\bar{p}_{\bm{v}^*})$.

@@

\clearpage
\bibliographystyle{rss}
\bibliography{paper}
